---
title: "Homework 2"
author: "Yuyi Lin"
date: "2026-10-04"
output: github_document
---

```{r load_libraries}
library(tidyverse)
library(readxl)
```
# Problem 1

```{r}
transit_df = 
  read_csv(
    "data/NYC_Transit.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")
  )|>
  janitor::clean_names()|>
  select(line, station_name, station_latitude, station_longitude, entry, vending, entrance_type, ada, starts_with("route")
  )|>
 mutate(
    entry =
      case_match(
        entry,
        "YES" ~ TRUE,
        "NO" ~ FALSE
      )
  )|>
  distinct(station_name, .keep_all=TRUE)

transit_df
  head(transit_df,20)
  tail(transit_df,5)
```
  
**Read and cleaned data. Retained 19 variables. Converted the entry variable from character (YES vs NO) to a logical variable using case_match function. Reduced redundancy and organized rows by 1 unique station name per row.**

This dataset includes several character and numerical variables describing the MTA train station. There are `r ncol(transit_df)` variables and `r nrow(transit_df)` observations total. This means that after reducing to 1 station name per row, we find that there are `r nrow(transit_df)` unique stations total. The data is not tidy, as route number and route name are not distinct variables.
  
```{r}
transit_df|>
  filter(ada==TRUE)|>
  select(station_name, line)|>
  distinct()
```

84 stations are ADA compliant.

```{r}
transit_df|>
  filter(vending =="NO")|>
  pull(entry)|>
  mean(na.rm=TRUE)
```

37.7% of station entrances/edits without vending allow entrance.

```{r}
transit_tidy_df = 
  transit_df|>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name")

distinct_a = 
  transit_tidy_df |>
  filter(route_name=="A")|>
  distinct(station_name, line)

distinct_ada_a = 
  transit_tidy_df |>
  filter(route_name=="A", ada==TRUE)|>
  distinct(station_name, line)
```

There are `r nrow(distinct_a)` distinct stations that serve the A train. Of the stations that serve the A, `r nrow(distinct_ada_a)` are ADA compliant.

# Problem 2

```{r}
trash_df = 
  read_excel("data/Trash_Wheel_2026.xlsx",
     na = c(""),
  sheet = "Mr. Trash Wheel",
  range = "A2:N743"
  )|>
  filter(!is.na(Dumpster))|>
  janitor::clean_names()|>
 mutate(
   year=as.numeric(year),
  plastic_bottles = as.numeric(plastic_bottles)
  )
  
professor_df = 
  readxl::read_excel("data/Trash_Wheel_2026.xlsx",
    na = c(""),
  sheet = "Professor Trash Wheel",
  range = "A2:M140"
  )|>
  filter(!is.na(Dumpster))|>
  janitor::clean_names()|>
 mutate(
   year=as.numeric(year),
  plastic_bottles = as.numeric(plastic_bottles)
  )

gwynnda_df = 
  readxl::read_excel("data/Trash_Wheel_2026.xlsx",
    na = c(""),
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:N395"
  )|>
  filter(!is.na(Dumpster))|>
  janitor::clean_names()|>
 mutate(
   year=as.numeric(year),
  plastic_bottles = as.numeric(plastic_bottles)
  )

merge_df = 
  bind_rows (trash_df, professor_df, gwynnda_df)|>
   mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

The new dataset, which ranges from `r merge_df|>pull(year)|>min(na.rm=TRUE)` to `r merge_df|>pull(year)|>max(na.rm=TRUE)`, has `r nrow(merge_df)` observations and `r ncol(merge_df)` variables, where the average number of sports balls collected was `r merge_df|>pull(sports_balls)|>mean(na.rm=TRUE)`. Key variables include which dumpster collected the trash, the time (month, year, and date) trash was collected, the weight and volume of trash, the number of different trash types (ex. plastic, cigarettes, sports balls), and the amount of electricity generated after burning. 

For available data, Professor Trash Wheel collected a total of `r professor_df|>pull(weight_tons)|>sum(na.rm=TRUE)` tons of trash.

```{r}
gwynnda_cig = 
  gwynnda_df |>
  filter (year==2022, month == "June") |>
  pull(cigarette_butts)|>
  sum(na.rm=TRUE)
```

Gwynnda collected a total of `r gwynnda_cig` cigarette butts in June 2022.

# Problem 3
 
```{r}
unzip("data/data_mci.zip",
 exdir = "data"
)

baseline_df = 
  read_csv("data/data_mci/mci_baseline.csv",
 skip = 1,
 na = c("$",".")
  )|>
  janitor::clean_names()|>
  mutate(
    sex = as.factor(sex),
    apoe4 = as.factor(apoe4)
  )|>
  filter(
    age_at_onset > current_age | is.na(age_at_onset)
  )
```

Originally, I had used the unzip and exdir function to put the new files in my data folder. After OH, Gustavo suggested I just clicked manually on the zip file, which immediately opened up the data_mci folder within my data folder. Rows (total=`r nrow(baseline_df)` observations) were participant ID number, and columns were participant characteristics (total=`r ncol(baseline_df)`variables) such as age at baseline, sex, carrier status, sex, and age at onset of MCI, if applicable.

Of the `r nrow(baseline_df)` participants recruited `r baseline_df |> filter(is.na(age_at_onset) == FALSE) |> nrow()` developed MCI. The average baseline age is `r baseline_df |> pull(current_age) |> mean (na.rm = TRUE)`, and `r baseline_df|>filter(sex=="0" & apoe4=="1")|>mean(na.rm = TRUE)` of women in the study were APOE4 carriers.

```{r}
amyloid_df = 
  read_csv("data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c("NA","","Na"))|>
  janitor::clean_names()

merge_mci_df = 
  inner_join (baseline_df, amyloid_df, by = c("id"="study_id")
  )

write_csv(
  merge_mci_df, ("data/merge_mci.csv")
  )
```

**Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.**
Dataset followed similar tidying steps as baseline_df where first column was skipped, missing codes were specified and removed, and text grammar was corrected. Similar to baseline_df, rows (total=`r nrow(amyloid_df)` observations) were participant ID. However, columns contained the different times (total=`r ncol(amyloid_df)` variables; 2, 4, 6, and 8 years) that have elapsed since the study baseline to the visit when biomarker amyloid was measured.

**Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.**
`r nrow(baseline_df) - nrow(merge_mci_df)` participants appeared in only the baseline dataset,and `r nrow(amyloid_df) - nrow(merge_mci_df)` participants appeared in only the amyloid dataset. The combined dataset now has `r nrow(merge_mci_df)` observations and `r ncol(merge_mci_df)` variables total.
!!!!!!!!!!

