---
title: "p8105_hw2_ysl2122"
author: "Yuyi Lin"
date: "2026-10-04"
output: github_document
---

# Problem 1

```{r}
library(tidyverse)
library(readxl)

# Read and clean data:
Transit_df = 
  read_csv(
    here::here("p8105_hw2_ysl2122", "Data","NYC_Transit.csv") 
  )|>
  janitor::clean_names()

Transit_df
  head(Transit_df,20)
  tail(Transit_df,5)
  skimr::skim(Transit_df)
```


```{r}
# Retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. 
# Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful)
Transit_df =
  Transit_df|>
  select(line, station_name, station_latitude, station_longitude, entry, vending, entrance_type, ada, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11)|>
  mutate(
    entry =
      case_match(
        entry,
        "YES" ~ 1,
        "NO" ~ 0
      )
  )|>
  distinct(station_name, .keep_all=TRUE)
  
Transit_df|>
  pull(ada)|>
  sum(na.rm=TRUE)

Transit_df|>
  filter(vending =="NO")|>
  pull(entry)|>
  mean(na.rm=TRUE)

Transit_tidy_df = 
  Transit_df|>
  mutate(
    across(route1:route11, as.character)
  )|>
    pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )

Transit_tidy_df|>
  filter(route_name=="A")|>
  distinct(station_name)|>
  nrow()

Transit_tidy_df|>
  filter(route_name="A")|>
  distinct(station_name, .keep_all=TRUE)|>
  filter(ada==TRUE)|>
  nrow())
```

#### Write a short paragraph about this dataset 
This dataset includes several character and numerical variables describing train routes, associated stations, and station details, such as station name, location (ex. location, longitude, latitude), and amenities/logistics (ex. vending, entrance type, ADA compliance, staffing capacity). / station location (station) – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

## Answer the following questions using these data:

**How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.**
  365 distinct stations

**How many stations are ADA compliant?**
  468 stations are ADA compliant

**What proportion of station entrances / exits without vending allow entrance?**
55.56% of stations

**Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?**
pending for OH!!!!!!!!

# Problem 2

```{r}
Trash_df = 
  read_excel(
    here::here("p8105_hw2_ysl2122", "Data", "Trash_Wheel.xlsx"),
  sheet = "Mr. Trash Wheel",
  range = "A2:N710",
  )|>
  mutate(Year=as.numeric(Year))
  
Professor_df = 
  readxl::read_excel(
    here::here("p8105_hw2_ysl2122", "Data", "Trash_Wheel.xlsx"),
  sheet = "Professor Trash Wheel",
  range = "A2:M135",
  )

Gwynnda_df = 
  readxl::read_excel(
    here::here("p8105_hw2_ysl2122", "Data", "Trash_Wheel.xlsx"),
  sheet = "Gwynns Falls Trash Wheel",
  range = "A2:N710",
  )|>
 mutate(Year=as.numeric(Year))

Merge_df = 
  bind_rows (Trash_df, Professor_df, Gwynnda_df)|>
  janitor::clean_names()|>
   mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

**Write a paragraph about these data** you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. 

**For available data, what was the total weight of trash collected by Professor Trash Wheel?** 

**What was the total number of cigarette butts collected by Gwynnda in June of 2022?**

# Problem 3
This problem uses data collected in an observational study to understand the trajectory of Alzheimer’s disease (AD) biomarkers. Study participants were free of Mild Cognitive Impairment (MCI), a stage between the expected cognitive decline of normal aging and the more serious decline of dementia, at the study baseline.

Basic demographic information were measured at the study baseline. The study monitored the development of MCI and recorded the age of MCI onset during the follow-up period, with the last visit marking the end of follow-up. APOE4 is a variant of the apolipoprotein E gene, significantly associated with a higher risk of developing Alzheimer’s disease. The amyloid 𝛽
 42/40 ratio holds significant promise for diagnosing and predicting disease outcomes. This ratio undergoes changes over time and has been linked to the manifestation of clinical symptoms of Alzheimer’s disease.
 
```{r}
unzip(
 here::here("p8105_hw2_ysl2122", "Data", "data_mci.zip"),
 exdir = here::here("p8105_hw2_ysl2122", "Data")
)

list.files(
  here::here("p8105_hw2_ysl2122", "Data"),
  recursive = TRUE
)

mci_df = 
  read_csv(
    here::here("p8105_hw2_ysl2122", "Data", "data_mci", "MCI_baseline.csv"),
 skip = 1
  )|>
  janitor::clean_names()|>
  mutate(
    sex = as.factor(sex),
    apoe4 = as.factor(apoe4)
  )

```
 and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.

